Lowestcase and Uppestcase letters: Further adventures in Derp Learning

### 2021-03-13 trunk/lowercase/paper/lowercase.tex
Just realized that I was writing about an older and much worse version of the vector net, ugh

\documentclass[twocolumn]{amsart}
\usepackage[top=0.75in, left=0.65in, right=0.65in, bottom=0.6in]{geometry}

\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}
% IPA symbols. safe turns off overrides for like \! which I still want
\usepackage[safe]{tipa}

@@

%     leftrule=0.2pt,rightrule=0.2pt,toprule=0.2pt,bottomrule=0.2pt,
%     enhanced jigsaw,
%    borderline horizontal={0.2pt}{0pt}{dashed},
%    borderline vertical={0.2pt}{0pt}{dashed}

@@

\begin{figure}[ht]
\includegraphics[width=0.5 \linewidth]{rotatej}
\caption{ Example alignment (from debug UI) of predicted shape (blue)
  to expected shape (green). We require each point to be mapped (red)
  to a point from the expected contour in a monotonic order (but
  several can be mapped to the same one), so that we can attribute
  error to each point. I hope it was fun to write this code because
  it was too slow to do pretty much anything but generate this figure!
} \label{fig:rotatej}
\end{figure}

% XXX add downloadable font too
\begin{figure*}[tp]
\includegraphics[width=0.9 \linewidth]{futurda}
\caption{ Type specimen for the generated font {\bf Futurda}. This is the
  classic font Futura, run through the vector based lowercasing model
  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
  \letterform{A}--\letterform{Z} run through the model and so ``made
  lowercase'' (it's obviously garbage). Next are
  \letterform{a}--\letterform{z},
  made even more lowercase. Also rubbish.
  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
  and ``Sphinx of black quartz, judge my vow!'' Although the text is
  illegible and kind of uninteresting, it does have a certain dynamic
  Rorschach aesthetic, like a collection of delicate moths pinned
  to paperboard, that one could consider framing or publishing in
  the proceedings of SIGBOVIK 2021.
} \label{fig:futurda}
\end{figure*}

@@

SDFs seem well-suited for machine learning. They contain more
information per pixel than a plain bitmap, so we can use a smaller
input and output size. On the input
side, extremal pixels that would almost never be set in a bitmap still
have significant information (distance to the character). The error
function is just pixel-by-pixel difference. The rendering of the
output is inherently tolerant of some noise because of the sampling
and thresholding. So, this seemed like it might work well! (It doesn't
work that well.)

I got some stats on the font database, and determined the following
parameters for the SDFs.
 (here, $36\times36$ pixel images)
% Figure: bitmap vs SDF

@@

@@

@@

@@

\nocite{murphy2019blind}
\nocite{murphy2019eloworld}

>>>>

\documentclass[twocolumn]{article} % {amsart}
\usepackage[top=0.75in, left=0.65in, right=0.65in, bottom=0.6in]{geometry}

\usepackage{float}
\usepackage{url}
% \usepackage{code}
% \usepackage{cite}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{chessboard}
% IPA symbols. safe turns off overrides for like \! which I still want
\usepackage[safe]{tipa}

@@

@@

\begin{figure}[ht]
\begin{centering}
  \includegraphics[width=0.5 \linewidth]{rotatej}
\end{centering}
\caption{ Example alignment (from debug UI) of predicted shape (blue)
  to expected shape (green). We require each point to be mapped (red)
  to a point from the expected contour in a monotonic order (but
  several can be mapped to the same one), so that we can attribute
  error to each point. I hope it was fun to write this code because
  it was too slow to do pretty much anything but generate this figure!
} \label{fig:rotatej}
\end{figure}

% FIXME: Wait I had much better screenshots, see video/lowercase

% XXX add downloadable font too
\begin{figure*}[tp]
\begin{centering}
  \includegraphics[width=0.9 \linewidth]{futurda}
\end{centering}
\caption{ Type specimen for the generated font {\bf Futurda}. This is the
  classic font Futura, run through the vector based lowercasing model
  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
  \letterform{A}--\letterform{Z} run through the model and so ``made
  lowercase'' (it's obviously garbage). Next are
  \letterform{a}--\letterform{z},
  made even more lowercase. Also rubbish.
  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
  and ``Sphinx of black quartz, judge my vow!'' Although the text is
  illegible and kind of uninteresting, it does have a certain dynamic
  Rorschach aesthetic, like a collection of delicate moths pinned
  to paperboard, that one could consider framing or publishing in
  the proceedings of SIGBOVIK 2021.
} \label{fig:futurda}
\end{figure*}

@@

\begin{figure}[ht]
  \includegraphics[width=0.95 \linewidth]{sdf-figure}
  \caption{
    The signed distance function representation of a letterform.
    At the very left, a $36\times 36$ pixel rasterization of
    the character without anti-aliasing, for comparison. Any
    scaling of this will have chunky pixel artifacts. Next,
    a $36\times 36$ pixel SDF of same. Third, simply scaling
    that $36\times 36$ image to $180 \times 180$ pixels with
    bilinear sampling. Finally, that image thresholded to
    produce a $180 \times 180$ pixel rasterization, which is
    far superior despite being derived from a $36 \times 36$ pixel
    bitmap. Typically this process is performed at an even higher
    scale and then downsampled to produce an anti-aliased image.
    } \label{fig:sdf}
\end{figure}

SDFs seem well-suited for machine learning. They contain more
information per pixel than a plain bitmap, so we can use a smaller
input and output size. On the input side, extremal pixels that would
almost never be set in a bitmap still have significant information
(distance to the character). The error function is just pixel-by-pixel
difference. The rendering of the output is inherently tolerant of some
noise because of the sampling and thresholding. So, this seemed like
it might work really well! (It doesn't work that well.)

I computed some stats on the font database, and determined the
following parameters for the fixed-size SDFs we train on. The images
are $36 \times 36$ pixels. The character box is placed such that there
are $2$ pixels of top padding, and $9$ pixels of left and bottom
padding. The character box is only ``nominal'' in the sense that the
font's contours can exceed its bounds, and this is completely normal
for a letter like \letterform{j} (which goes below the baseline and
often hangs to the left of the origin as well). I used an ``on-edge
value'' of 220 (because much more of the SDF is outside the letter
than inside) and the distance is scaled as 15 units per pixel (so that
pixels on the outer edge often have non-zero values). Compared to the
first version, I was somewhat more permissive in what fonts I trained
on, since there was no inherent limit to the number of contours their
complexity. I did exclude fonts whose rasterizations exceeded the
bounds of the SDF, which is possible (very wide \letterform{W} or
low-descending \letterform{j} perhaps) but rare.

@@

Having committed to the representation, again it is ``just'' a matter
of heating up the GPU to apply some linear and non-linear transforms.
The initial network had an input size of $36 \times 36 = 1296$ for the
SDF, and the output the same plus $26$ bonus outputs (one for each
letter, as before). I started with three hidden layers of $1296$, $1296$,
and $2916$ nodes, each sparse ($80\%$ of the weights are zero). Again,
don't take notes. This one works a bit better than before, but still
not impressive. The node references are assigned spatially (something
like the $20\%$ of the nodes on the previous layer that are closest to
the next layer's node) but due to a bug the spatial locality is
actually pretty strange.
% XXX if I show layer weights, forward-reference them here:
% It's responsible for the XXX
Every layer's transfer function is ``leaky relu'' ({\tt x > 0 \? x : 0.1 * x})
which is fast to compute and better than sigmoids in the output since
correct values will not just be 0 and 1. It would definitely make sense
to use convolutional layers for this problem, as features like serifs,
lines, curves, and so on could appear throughout the input and output.
I just haven't built support for that in my weird home-grown software,
yet.

I also adapted my weird home-grown software to train the
\makeuppercase\ and \makelowercase\ models simultanously. Two models
fit comfortably in GPU memory, with plenty of space for a stream of
training data (one training instance is only about 10kb). The only
challenging thing is arranging for them to feed each other generated
``inversion'' examples (Figure~\ref{fig:cotraining}), but this is just
a matter of programming, thank God. I should remember to do projects
that are mostly a matter of programming. Each round, $25\%$ of the
batch consists of inverted examples from the symmetric model's output
from a recent round. I make sure that one model is not allowed to
get too far ahead (more than 2 rounds) of the other in training,
because I want this feedback loop to be somewhat tight.

So I did that and let it run for a month. Actually I had to start over
several times with different parameters and initialization weights
because it would get stuck

% you propagate my back, I'll propagate yours

@@

% TODO loss over time

% some other things that are worth a mention:
%   autoparallel
%   efficiency improvements (restrict, compile-time constants)

@@

lorem ipsum (Figure~\ref{fig:perfect})

\begin{figure*}[tp]
\begin{centering}
  \includegraphics[width=0.9 \linewidth]{perfecthallucination}
\end{centering}
\caption{
%  XXX Why doesn't this caption come out the way I'm expecting?
%  (in amsart style... can't have two figure*s on the same page?)
   Type specimen for the generated font {\bf Perfect Hallucination}.
  TODO write a description!
} \label{fig:perfect}
\end{figure*}

% LSDF

dolor sit amet

\section{Chess-playing}

@@

\section{Acknowledgements}

For this project I used \verb+stb_truetype.h+ from the excellent stb
collection of single-file libraries.~\cite{stb} I did push its limits
somewhat and ``automatically discovered'' assertion failures and other
crashes (e.g. during blackbox optimization on all 100k fonts), but it
saved a lot of time. This library only helps with reading the fonts
and generating SDFs. To generate TTFs, I had to write my own pipeline,
which generates FontForge's .SFD (a typo factory right there) files,
and then did the final export with FontForge~\cite{fontforge}. Thanks
Jason Reed for this suggestion.

\nocite{murphy2019blind}
\nocite{murphy2019eloworld}
\nocite{stb}
